% =================================================================
% CHAPTER 5: DISCUSSION
% =================================================================
\chapter{Discussion}
\label{ch:discussion}

This chapter interprets the findings reported in Chapter~\ref{ch:experiments}. Rather than repeating the quantitative tables, it explains why the proposed two-stage framework behaves as observed, identifies the principal architectural limitations, and discusses the implications of these findings for semantically controlled video generation in autonomous driving.

\FloatBarrier
% -----------------------------------------------------------------
\section{Analysis of Experimental Results}
\label{sec:result_analysis}

\subsection{Semantic VAE Performance}
\label{subsec:vae_analysis}

The Semantic VAE results in Section~\ref{sec:vae_results} show that a semantic-native latent representation is essential for this thesis. The strongest finding is not only the final 89.7\% mIoU, but the large gap between the RGB baseline and the semantic-native design. Once the 3-channel RGB bottleneck is removed, the encoder no longer has to force discrete trainID boundaries through a latent space that was pretrained for natural-image appearance. The resulting improvement demonstrates that semantic segmentation maps should be treated as categorical structure rather than as ordinary RGB imagery.

The loss ablations clarify why the final architecture is effective. Boundary weighting produces the largest single improvement because semantic errors are concentrated at class transitions, where even a one-pixel drift can change the label identity of thin structures or object contours. Dice loss makes a smaller absolute contribution, but it remains important because it counteracts the strong class imbalance of KITTI-360. These results together indicate that reconstruction quality in semantic latent space is governed by both representation design and objective design. A suitable latent architecture alone is not sufficient if the optimisation signal still under-weights boundaries and rare classes.

The main implication for the remainder of the thesis is that the Semantic VAE is not the dominant failure point of the overall framework. It provides a sufficiently strong semantic representation to make Stage~1 prediction and Stage~2 rendering meaningful research problems. The larger limitations therefore arise in how this latent representation is used by the diffusion stages rather than in whether semantic maps can be encoded at all.

\subsection{Stage~1 Semantic Prediction}
\label{subsec:stage1_discussion}

The Stage~1 results in Section~\ref{sec:stage1_results} show that future semantic prediction from a single initial RGB frame remains substantially harder than semantic reconstruction. The model reaches 40.01\% mIoU on 170 validation clips, with strong performance on large and mostly static classes such as road, building, sky, vegetation, truck, and car, but markedly weaker performance on thin or rare classes such as pole, rider, bicycle, and traffic light. This class pattern indicates that the model captures coarse scene topology reliably, yet does not preserve balanced semantic structure across all categories.

This behaviour is consistent with the conditioning design of Stage~1. The diffusion model is strongly anchored to the initial semantic layout and is additionally evaluated in a boundary-conditioned setting, where the final semantic frame is also supplied during denoising. Under this formulation, the simplest solution is to preserve the initial arrangement of large regions and interpolate smoothly between anchors. Such a bias is beneficial for static road layout and background geometry, but it is unfavourable for newly appearing agents, thin structures, or temporally local events. Stage~1 therefore behaves more like a structured scene-persistence model than a fully open-ended semantic forecaster.

The interpretation of the Stage~1 score must follow this distinction. The reported mIoU is informative, but it does not correspond to a pure frame-0-only forecasting problem. It measures semantic sequence generation under strong structural guidance. This matters for the thesis discussion because the current limitation is not simply insufficient optimisation. The more fundamental issue is that the conditioning and representation biases encourage persistence of dominant structure while leaving minority classes and motion-dependent changes weakly constrained.

\subsection{Stage~2 Semantic-to-Video Generation}
\label{subsec:stage2_discussion}

The Stage~2 results in Section~\ref{sec:stage2_results} confirm that semantic segmentation maps can guide photorealistic video generation, and they demonstrate competitive semantic alignment with respect to reference methods. The most important architectural finding is the effect of partial UNet unfreezing. Allowing the bottleneck and output projection of the Stable Video Diffusion backbone to adapt raises DRN-based semantic fidelity from 23.20\% to 50.64\% mIoU — evaluated using the same multi-scale, confidence-weighted DRN-D-105 protocol as the reference benchmark — while increasing pixel accuracy to 85.34\% and frequency-weighted IoU to 75.31\%. This gain is too large to interpret as a minor optimisation effect. It shows that the frozen decoder path was previously too rigid to translate ControlNet features into semantically faithful RGB frames.

However, the per-class Stage~2 results reveal that the improvement is still concentrated on large contiguous regions. Car (92.17\%), road (89.81\%), building (86.40\%), vegetation (86.20\%), and sky (85.85\%) all exceed 85\% IoU, whereas truck and bus reach 0.00\% and person achieves only 14.93\%. The generated video is therefore semantically faithful at coarse scene scale but not yet at the level of fine structure or rare objects. The high pixel accuracy together with the lower mIoU has the same interpretation as in Stage~1: dominant classes occupy most pixels, so the model appears semantically strong in aggregate while still omitting visually small but semantically important categories.

The multi-scale reinjection experiment sharpens this interpretation. Reinjection improves the realism-oriented metrics, reducing FID from 21.9122 to 20.9585 and FVD from 255.2064 to 234.3064, but it lowers DRN-based semantic fidelity from 50.64\% to 43.69\% mIoU under the same evaluation protocol. The semantic cost is systematic: the most plausible reading is that the current reinjection design reinforces global layout and temporal smoothness more effectively than it preserves fine semantic boundaries.

Qualitative inspection of the lowest-scoring validation clips supports the same conclusion. The weakest Stage~2 results are concentrated in sequences characterised by abrupt viewpoint change, junction transitions, and vegetation-dominated scenes, whereas performance is stronger in clips with more regular forward motion and stable road layout. This concentration indicates that Stage~2 does not fail uniformly across the validation set. Instead, semantic control becomes less reliable when the scene undergoes rapid structural change, which is consistent with the observed reductions in fidelity during turns and complex intersections.

Comparison with external reference methods sharpens the same interpretation. When evaluated under the same DRN-D-105 multi-scale protocol, the UNet-unfreeze model achieves 50.64\% mIoU, surpassing SVS-GAN~\cite{seyam2024svsgan} (45.08\% mIoU), while also yielding markedly lower FVD. This indicates that the proposed diffusion-based Stage~2 achieves both stronger semantic alignment and better temporal realism than the autoregressive GAN baseline on KITTI-360, as measured by this evaluation protocol. The comparison with Ctrl-V~\cite{luo2025ctrlv} is methodological rather than benchmark-level, because Ctrl-V uses sparse bounding-box control on KITTI rather than dense semantic control on KITTI-360. Nevertheless, it clarifies the central challenge addressed in this thesis: replacing sparse object-level control with dense semantic control increases scene-level specificity, but it also makes faithful rendering harder because the generator must preserve structure across the entire image rather than only around selected objects.

\subsection{Cross-Stage Interpretation}
\label{subsec:cross_stage_discussion}

The relationship between Stage~1 and Stage~2 semantic fidelity requires careful interpretation because the two components are assessed under different evaluation protocols. Stage~1 reaches 40.01\% mIoU on direct semantic prediction (170 clips, 4{,}250 frames, single-scale DRN evaluation), while the best Stage~2 configuration reaches 50.64\% mIoU when semantic fidelity is measured by re-segmenting generated RGB frames using the multi-scale, confidence-weighted DRN protocol on 475 specifically selected frames. The higher Stage~2 score under this protocol is partly attributable to the more focused evaluation set — 19 clip groups chosen to match the reference benchmark — and to the stronger multi-scale inference, which benefits large and well-textured classes disproportionately. Direct numerical comparison between the two figures is therefore not straightforward.

Regardless of this protocol difference, both stages contribute semantic loss that limits the performance of the full end-to-end system. Stage~2 is evaluated under an idealised setting in which ground-truth semantic maps are supplied as conditioning input; in the full pipeline, Stage~2 must render from Stage~1 predictions rather than from ground truth, and any semantic errors introduced in Stage~1 propagate through to the RGB generation. Improving only one stage will therefore not remove the principal bottleneck of the overall system. A better Stage~1 predictor reduces semantic noise at the renderer input, while a better Stage~2 renderer preserves more of the semantic structure already present in the control signal. The two stages therefore remain coupled research problems even though they are trained independently.

\FloatBarrier
% -----------------------------------------------------------------
\section{Architectural Limitations of the Two-Stage System}
\label{sec:stage1_limitations}

This section discusses the principal architectural factors that explain the result patterns described above.

\subsection{Stage~1 Limitations}
\label{subsec:stage1_limitations_detail}

The main Stage~1 limitation is a strong copy-forward bias. The diffusion model receives a repeated semantic representation of the first frame across the intermediate temporal slots and an additional semantic anchor at the end of the clip. This conditioning is useful for stabilising coarse scene layout, but it favours semantic persistence over semantic change. As a result, the model preserves road, building, sky, and vegetation effectively while under-reacting to classes whose future state diverges from the initial frame. The present formulation therefore makes interpolation between structural anchors easier than genuine forecasting of unseen intermediate semantics.

The second Stage~1 limitation is the combination of aggressive spatial compression and missing motion information. Semantic maps are encoded at an effective latent resolution of $24 \times 88$, which already removes much of the spatial support available to thin or small classes. At the same time, the model receives only a single initial RGB frame as visual evidence and does not explicitly observe ego-motion, optical flow, or depth. Future semantics in a driving scene depend strongly on these cues because turning, occlusion, and object motion all alter which structures remain visible. The current architecture is therefore asked to infer temporally evolving categorical structure from a representation that is both spatially compressed and only weakly motion-aware.

The Stage~1 analysis also suggests that optimisation is not perfectly aligned with the semantic latent distribution. The semantic latents are injected into a diffusion backbone whose original noise regime was learned for RGB latent statistics, and the training objective operates in continuous latent space rather than directly on class-balanced semantic outputs. This combination is sufficient for coarse layout learning, but it offers comparatively weak protection for rare classes and precise boundaries. The remaining Stage~1 gap is therefore best understood as an information-flow limitation rather than as a simple shortage of trainable parameters.

\subsection{Stage~2 Limitations}
\label{subsec:stage2_limitations_detail}

The main Stage~2 limitation is that semantic control is spatially strong but temporally weak. The semantic sequence enters ControlNet through a frame-wise 2D conditioning pathway, while explicit temporal mixing occurs only later inside the shared spatio-temporal blocks of the pretrained video backbone. This means that semantic control is not temporally structured at entry. It is instead expected to become temporally coherent after fusion with a backbone that was originally trained for natural RGB video generation. This design is adequate for smooth clips, but it is less effective when viewpoint changes rapidly or when large parts of the scene enter and leave view.

A second Stage~2 limitation is the persistent appearance anchor created by repeating the first RGB latent across all generated frames. This conditioning is beneficial for preserving scene identity and texture, but it also introduces appearance inertia. When the future semantic map requires large geometric change, the renderer receives two competing signals: the semantic control suggests a new spatial arrangement, while the repeated first-frame RGB latent encourages the decoder to remain close to the initial view. This conflict is particularly relevant in clips with turns, junction transitions, or strong foreground-background motion.

The current Stage~2 architecture also remains constrained by the same $8\times$ semantic compression used in Stage~1. Semantic latents start at $24 \times 88$ and are further processed through deeper bottlenecks at $12 \times 44$, $6 \times 22$, and $3 \times 11$. Multi-scale reinjection partly refreshes semantic information, but it does so at increasingly coarse scales and only through the encoder path. This is sufficient to reinforce global region identity, yet it is not well matched to preserving thin structures or sharp class boundaries. The reinjection result (43.69\% vs.\ 50.64\% mIoU) therefore does not reject multi-scale conditioning as a concept. It shows that the present form of coarse reinjection improves realism more readily than it improves fine-grained semantic control.

The final Stage~2 limitation concerns both training objective and evaluation protocol. The model is trained primarily with a latent denoising objective in RGB latent space, which encourages plausible appearance but does not directly penalise the disappearance of small semantic classes. The reported Stage~2 mIoU is measured through an external DRN segmenter under ground-truth semantic conditioning and on a fixed set of 475 frames covering 15 of 19 semantic classes. This remains a useful controllability metric, but it is an upper bound on full pipeline performance and reflects only the classes present in the 19 selected clip groups. The current Stage~2 discussion must therefore distinguish carefully between semantic fidelity under ideal control and semantic robustness in the actual end-to-end setting.

\FloatBarrier
% -----------------------------------------------------------------
\section{Implications for Future Design}
\label{sec:stage1_future}

The combined evidence from Stage~1 and Stage~2 points to a common research priority: the next gains are more likely to come from improving information flow than from simply enlarging the denoising backbone. In Stage~1, this means reducing the dependence on repeated semantic anchors and supplying stronger motion-related cues, such as multiple RGB frames, ego-motion, depth, or optical flow. In Stage~2, it means replacing the frame-wise semantic entry path with a genuinely spatio-temporal semantic conditioner so that semantic control remains coherent in clips with abrupt viewpoint change.

The second common priority is higher-resolution semantic guidance. Both stages currently operate on the same $24 \times 88$ semantic latent grid, which is sufficient for global scene layout but not for reliable preservation of thin or rare classes. A higher-resolution or hierarchical semantic representation would therefore benefit the full pipeline. For Stage~2 in particular, future variants could combine coarse ControlNet-style residual guidance with high-resolution adapter pathways and stronger temporal conditioning, following the general direction of controllable diffusion architectures such as ControlNet, T2I-Adapter, and video-conditioned control models~\cite{zhang2023adding,mou2024t2i,hu2023videocontrolnet}.

The third priority is better alignment between evaluation and the intended deployment setting. Stage~1 should ultimately be reported in a strict frame-0-only setting in addition to the current boundary-conditioned protocol, while Stage~2 should be trained and evaluated not only with ground-truth semantic maps but also with the imperfect semantic predictions produced by Stage~1. This would make the controllability claim of the full system more defensible because it would measure the actual robustness of the two-stage interface rather than only the upper bound of each component in isolation.

\FloatBarrier
% -----------------------------------------------------------------
\section{Chapter Summary}
\label{sec:conclusion}

The discussion in this chapter shows that the thesis contributions are meaningful, but also that the current framework remains limited by strong persistence biases and coarse semantic control. The Semantic VAE successfully establishes a compatible latent interface for semantic maps, Stage~1 learns future scene topology but not balanced semantic forecasting, and Stage~2 translates semantic control into realistic RGB video but still loses fine structure, especially under temporally difficult scene regimes. The central conclusion is therefore that semantic control is feasible, yet not fully robust at the current spatial resolution and conditioning design. Chapter~\ref{ch:summary} summarises these contributions and outlines the most important next research directions.
